\chapter{Context \& Problem Formulation} \label{chap:problem-formulation}

\begin{chapquote}{\textit{Judea Pearl}}
``As X-rays are to the surgeon, graphs are for causation.''
\end{chapquote}

The main goal of this chapter is to underpin the essential theoretical background and terminology needed to formally define our work:  Causal discovery from time-series data using a foundational neural approach (our LCMs). The purpose of this part is to serve as a conceptual and notational bridge between the high-level motivation discussed in Chapter \ref{chap:introduction} and the following chapters. We focus on Temporal Structural Causal Models (TSCMs) and key assumptions needed for principled causal discovery. Rather than providing a complete treatment of the theory of Causality, our aim is to introduce the necessary formalism tailored to our own setting. For readers seeking a broader perspective on causality, we refer to the seminal works by \citet{pearl2009causality, pearl2018book}, \citet{spirtes2001causation} and \citet{peters2017elements}. For the case of causality and causal discovery in time-series, we point the reader to the works of \citet{runge2018causal, runge2023causal, assaad2022survey, gong2024causal}. We also elaborate on some important aspects of deep learning, as it also serves as an integral part of our work. For foundational reading, we refer to the textbook by \citet{goodfellow2016deep}. Section \ref{sec:scms} introduces Structural Causal Models, which are then extended to Temporal SCMs (TSCMs) in Section \ref{sec:temp-scms}, while Section \ref{sec:causal-assumptions} outlines the assumptions necessary for causal identifiability and model interpretation. Section \ref{sec:dl} provides a short overview of the necessary deep learning terminology for the unfamiliar reader. Finally, Section \ref{sec:problem-formulation} defines the main objective of our work.

\input{chapters/background/causality/causality}
\input{chapters/background/deep-learning/deep-learning}
\input{chapters/background/problem/problem}