\chapter*{Preface}
\addcontentsline{toc}{chapter}{Preface}

\begin{chapquote}{Tao Te Ching, \textit{Lao Tzu}}
``A journey of a thousand miles begins with a single step.''
\end{chapquote}

This thesis marks the culmination of a long journey into the world of causality. The work presented was carried out in collaboration with the Foundation for Research and Technology Hellas in Heraklion and Huawei Ireland Research Centre in Dublin, Ireland. It represents part of a joint research project that brought together academic research and industrial perspectives on developing robust causal discovery methods under deep learning techniques, for real-world applications.

Writing this thesis has proven to be both a technical challenge and a personal journey, including multiple setbacks, experimental loops as well as positive surprises and breakthroughs. I hope that this work will not only contribute to advancing robust causal discovery methods for decision making, but also inspire further research in building machine learning systems that can reason about the world, instead of merely recognizing patterns.

\section*{Acknowledgements}

I would first like to thank my advisor for the guidance and pointers towards this work, on a novel line of research that was not without its own challenges and struggles. His vision and enthusiam has shaped much of this work. A special thanks goes to Dr. Nikolaos Gkorgkolis, as both a collaborator and postdoctoral mentor, whose insight and enthusiasm for scientific discovery have shaped much of this work. I am grateful for our thoughtful discussions, creative ideas, countless meetings, software engineering, research writing and planning. I also want to thank all the members of the AIOps team at Huawei Ireland Research Centre: Wang MingXue, who led this collaboration; Bora Caglayan, Andrea Tonon and Dario Simionato, for bridging the gap between the two teams with patience. Working with such talented people has been a rewarding aspect of this journey. Finally, I want to express my deepest gratitude to my family for their unwavering love and encouragement. This includes my mother Sofia, my father Michalis and my brother Christos. Nothing in my life would have been possible without them.

\section*{Publications}

This thesis is based upon the following papers:

\begin{enumerate}
    \item \textit{Temporal Causal-based Simulation for Realistic Time-Series Generation} \\
          Nikolaos Gkorgkolis$^*$, \underline{Nikolaos Kougioulis}$^*$, MingXue Wang, Bora Caglayan, Andrea Tonon, Dario Simionato \& Ioannis Tsamardinos \\
          (ArXiv Preprint: arXiv:2025.01.01)

    \item \textit{Large Causal Models for Temporal Causal Discovery} \\
          \underline{Nikolaos Kougioulis}, Nikolaos Gkorgkolis, MingXue Wang, Bora Caglayan, Andrea Tonon, Dario Simionato \& Ioannis Tsamardinos \\
          (In preparation)
\end{enumerate}

\def\thefootnote{*}\footnotetext{Authors contributed equally to this work.}\def\thefootnote{\arabic{footnote}}

%The following works, although not an integral work of this thesis, were also completed during its preparation.

%\begin{enumerate}
%    \item ...
%\end{enumerate}